---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r, echo=TRUE, message=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(dplyr)
```
## Loading and preprocessing the data
```{r, echo=TRUE, message=FALSE}
data <- read.csv(unz("activity.zip", "activity.csv"))
data$date <- ymd(data$date ,tz = "America/New_York") 
```
## What is mean total number of steps taken per day?
```{r, echo=TRUE, message=FALSE}
stepsPerDay <- data[!is.na(data$steps),] %>% group_by(date) %>% summarise(Sum = sum(steps))
hist(stepsPerDay$Sum, main = "Histogram of Daily Steps", xlab = "Daily Steps")
```

#### The mean is

```{r, echo=TRUE, message=FALSE}
mean(stepsPerDay$Sum)
```
#### The median is 
```{r, echo=TRUE, message=FALSE}
median(stepsPerDay$Sum)
```

## What is the average daily activity pattern?
```{r, echo=TRUE, message=FALSE}
stepsPerInterval <- data[!is.na(data$steps),] %>% group_by(interval) %>% summarise(Average = mean(steps))
qplot(interval, Average, data = stepsPerInterval, geom="line") 
```

#### the interval with the maximum steps is 
```{r, echo=TRUE, message=FALSE}
stepsPerInterval[stepsPerInterval$Average == max(stepsPerInterval$Average), ]
```
## Imputing missing values
```{r, echo=TRUE, message=FALSE}
length(data[is.na(data$steps),1])
dataCleaned <- data
```

#### replacing the NAs with the average for that interval
```{r, echo=TRUE, message=FALSE}
dataCleaned[is.na(dataCleaned$steps), 1] <- merge(dataCleaned[is.na(dataCleaned$steps), ], stepsPerInterval, by = "interval")$Average
stepsPerDayCleaned <- dataCleaned %>% group_by(date) %>% summarise(Sum = sum(steps))
hist(stepsPerDayCleaned$Sum, main = "Histogram of Daily Steps Cleaned", xlab = "Daily Steps")
```

#### The mean is

```{r, echo=TRUE, message=FALSE}
mean(stepsPerDayCleaned$Sum)
```

#### The median is 
```{r, echo=TRUE, message=FALSE}
median(stepsPerDayCleaned$Sum)
```

#### Note the median has shifted to the right as compared to the un-cleaned data, but the mean remained the same

## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE, message=FALSE}
dataCleaned$dayType <- as.factor(sapply(dataCleaned$date, function(x) if(wday(x) %in% 1:5) {"weekday"} else {"weekend"}))
stepsPerDayTypeCleaned <- dataCleaned %>% group_by(interval, dayType) %>% summarise(Average = mean(steps))
qplot(interval, Average, data = stepsPerDayTypeCleaned, geom="line", facets = dayType~., ylab = "average number of steps")

```